Identifying Effective Motivational Interviewing Communication Sequences Using Automated Pattern Analysis

### 2018-09-04 jhir/latex-manuscript/manuscript_publish_highlighted.tex
prepared final version of manuscript for publication

\date{Received: 3 December 2017 /Revised: 16 April 2018 and 29 June 2018 /Accepted: }
% The correct dates will be entered by the editor

@@

The sequential analysis procedure used in the above MI process studies~\cite{moyers2006therapist, catley2006adherence, thrasher2006motivational, mccambridge2011fidelity} is based on the first-order Markov Chain model~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}. The Markov Chain model is a discrete-time stochastic process built on the assumption that the state of a system or condition changes over time and only depends on the previous event. Hence, Markov Chain models have two main drawbacks. The first is their inability to preserve the long-range dependencies between observations in a sequence. However, in MI, an observed behavior can be influenced by any of the preceding behaviors. The second drawback is their inability to consider similarities between behavior codes and, consequently, their inability to identify multiple similar behaviors that lead to the same outcome. Thus, first-order Markov models may be insufficient to fully understand the associations between behaviors in patient-counselor communication sequences. Therefore, there is a need for more powerful computational methods, which consider clusters of similar behavior codes and long-range dependencies between behaviors, to identify causal relationships. The goal of the current research is to test the applicability of data mining and machine learning methods to identify effective patterns of patient-counselor communication.

@@

\rowcolor{lightgraycol}
\noalign{\smallskip}
\multicolumn{4}{l}{\textbf{Patient}} \\
\noalign{\smallskip}
CT & Change Talk & Statements that express the patient's desire, ability, reasons, need for, or commitment to (intentions, plans and action steps) changing their behavior & ``I will try to buy less junk food.''\\
ST & Sustain Talk & Statements that express the patient's desire, ability, reasons, need for, or commitment to (intentions, plans and action steps) to maintain the status quo or not change their behavior  & ``I didn't get to the gym this week.''\\
HUPW & High uptake weight & A turn that does develop the topic of the conversation. High Uptake statements include: weight related statements about actions of commitment, change talk and ambivalence that occurred in the past, patient questions to the counselor and session interruptions by persons who are not an active part of the treatment session. & ``Support is always good.  You know that's a key factor.  Mm-hmm.''\\
HUPO & High uptake other & An utterance that develops the topic of the conversation but is about non-target behaviors or interruptions & ``Yeah because the mentor comes and they take off and they go someplace for a little while.''\\
LUP & Low uptake  & An utterance that does not develop the topic of conversation but still allows it to continue & ``Mm-hmm.  Right.''\\
\hline
\end{longtable}
\end{small}

@@

\subsubsection{Frequent Pattern Mining}
\label{subsubsec:frequentPatternMining}
We applied frequent pattern mining to identify frequently occurring patterns of patient-counselor behavior codes in successful and unsuccessful communication sequences. Behavior codes in these patterns may be separated by one or more other codes. For this purpose, we utilized FPClose~\cite{grahne2005fast}, an efficient state-of-the-art closed frequent pattern mining algorithm implemented in SPMF~\cite{fournier2014spmf, fournier2016spmf}, to identify frequent patterns of patient-counselor communication behaviors in successful and unsuccessful MI communication sequences. SPMF is an open-source library providing more than 150 data mining algorithms. Popular non-closed frequent pattern mining algorithms include Apriori~\cite{agrawal1994fast} and FP-Growth~\cite{han2000mining}. A \textit{frequent pattern} is defined as a pattern of observations, which appears in a given set of sequences more often than a user-specified threshold called the \textit{minimum support count}. For example, \{A\}, \{C\}, \{D\} and \{C, D\} are frequent patterns in the example set of sequences in Figure~\ref{fig:fpm}, since these patterns appear at least 2 times, which is the minimum support count in this example. In this work, we identified and analyzed \textit{closed frequent patterns} among all sequences of behavior codes in successful and unsuccessful communication sequences. A \textit{frequent pattern} is \textit{closed} if none of its supersets have the same support count~\cite{pasquier1999discovering}, where a set X is a superset of another set Y, if X contains all the elements of the set Y. For example, the itemsets \{A\}, \{C\}, \{D\} and \{C, D\} in Figure~\ref{fig:fpm} are not closed frequent patterns since their supersets \{A, B\}, \{B, C\}, \{B, D\} and \{B, C, D\} have the same support count. Therefore, \{B\}, \{A, B\}, \{A, D\}, \{B, C\}, \{B, D\} and \{B, C, D\} are closed frequent patterns since none of their supersets have the same support count. On the other hand, itemsets \{A, C\}, \{A, B, C\}, \{A, B, D\}, \{A, C, D\} and \{A, B, C, D\} have support counts of 1, 0, 1, 0 and 1, respectively, which is less than the minimum support count and thus are identified as non-frequent itemsets. Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we followed prior work~\cite{moradi2017quantifying, liu1999mining} and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences. For each pattern, the statistical significance of the difference between pattern occurrence in successful and unsuccessful communications was reported by computing p-value of Pearson's chi-square test.

@@

This study is part of a line of research to develop machine-learning models to annotate (code) and analyze patient-counselor communication patterns. We have previously reported on the development of probabilistic generative models~\cite{kotov2014towards, kotov2015interpretable} and application of novel features for maximum margin and deep learning classifiers~\cite{hasan2016study} with the goal of automated annotation of MI session transcripts. Experiments applying the annotation model to novel datasets are underway to assess the generalizability of the model to more diverse types of clinical encounters (e.g, email coaching to increase fruit and vegetable intake, HIV clinical care visits~\cite{carcone2018using}). We also developed and evaluated probabilistic and deep learning methods for the task of predicting success of motivational interviews at any point during the interview~\cite{hasan2018predicting}. In this paper, we present current work that builds on past work performed on automation of annotation, more specifically, the sequential analysis of patient-counselor communication data to identify the counselor communication strategies linked to the elicitation of change talk and sustain talk. Next steps include examining the performance of these models in diverse data sets representing different populations and behavioral problems. We are also experimenting with machine learning methods to segment (parse) the stream of communications in a healthy eating promotion intervention delivered by email into semantically coherent chunks (phrases, sentences or paragraphs), to which the annotation model can then assign a behavior code. Together, these models form the basis of a complete system to automatically code and analyze patient-counselor interactions. An automated system for behavioral coding and analysis could substantially accelerate the pace of research on the causal mechanisms of Motivational Interviewing and inform both the theory and clinical practice by providing clinicians with information about how to best tailor their communication strategies to different patient populations.

>>>>

\date{Received: 3 December 2017 /Revised: 16 April 2018 and 29 June 2018 /Accepted: 31 August 2018}
% The correct dates will be entered by the editor

@@

The sequential analysis procedure used in the above MI process studies~\cite{moyers2006therapist, catley2006adherence, thrasher2006motivational, mccambridge2011fidelity} is based on the first-order Markov Chain model~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}. The Markov Chain model is a discrete-time stochastic process built on the assumption that the state of a system or condition changes over time and only depends on the previous event. Hence, Markov Chain models have two main drawbacks. The first is their inability to preserve the long-range dependencies between observations in a sequence. However, in MI, an observed behavior can be influenced by any of the preceding behaviors. The second drawback is their inability to consider \hl{the} similarities between behavior codes and, consequently, their inability to identify multiple similar behaviors that lead to the same outcome. Thus, first-order Markov models may be insufficient to fully understand the associations between behaviors in patient-counselor communication sequences. Therefore, there is a need for more powerful computational methods, which consider clusters of similar behavior codes and long-range dependencies between behaviors, to identify causal relationships. The goal of the current research is to test the applicability of data mining and machine learning methods to identify effective patterns of patient-counselor communication.

@@

\rowcolor{lightgraycol}
\noalign{\smallskip}
\multicolumn{4}{l}{\textbf{Patient}} \\
\noalign{\smallskip}
CT & Change Talk & Statements that express the patient's desire, ability, reasons, need for, or commitment to (intentions, plans and action steps) changing their behavior & ``I will try to buy less junk food.''\\
ST & Sustain Talk & Statements that express the patient's desire, ability, reasons, need for, or commitment to (intentions, plans and action steps) to maintain the status quo or not change their behavior  & ``I didn't get to the gym this week.''\\
HUPW & High uptake weight & A turn that does develop the topic of the conversation. High Uptake statements include: \hl{weight-related} statements about actions of commitment, change talk and ambivalence that occurred in the past, patient questions to the counselor and session interruptions by persons who are not an active part of the treatment session. & ``Support is always good.  You know that's a key factor.  Mm-hmm.''\\
HUPO & High uptake other & An utterance that develops the topic of the conversation but is about non-target behaviors or interruptions & ``Yeah because the mentor comes and they take off and they go someplace for a little while.''\\
LUP & Low uptake  & An utterance that does not develop the topic of conversation but still allows it to continue & ``Mm-hmm.  Right.''\\
\hline
\end{longtable}
\end{small}

@@

\subsubsection{Frequent Pattern Mining}
\label{subsubsec:frequentPatternMining}
We applied frequent pattern mining to identify frequently occurring patterns of patient-counselor behavior codes in successful and unsuccessful communication sequences. Behavior codes in these patterns may be separated by one or more other codes. For this purpose, we utilized FPClose~\cite{grahne2005fast}, an efficient state-of-the-art closed frequent pattern mining algorithm implemented in SPMF~\cite{fournier2014spmf, fournier2016spmf}, to identify frequent patterns of patient-counselor communication behaviors in successful and unsuccessful MI communication sequences. SPMF is an open-source library providing more than 150 data mining algorithms. Popular non-closed frequent pattern mining algorithms include Apriori~\cite{agrawal1994fast} and FP-Growth~\cite{han2000mining}. A \textit{frequent pattern} is defined as a pattern of observations, which appears in a given set of sequences more often than a user-specified threshold called the \textit{minimum support count}. For example, \{A\}, \{C\}, \{D\} and \{C, D\} are frequent patterns in the example set of sequences in Figure~\ref{fig:fpm}, since these patterns appear at least 2 times, which is the minimum support count in this example. In this work, we identified and analyzed \textit{closed frequent patterns} among all sequences of behavior codes in successful and unsuccessful communication sequences. A \textit{frequent pattern} is \textit{closed} if none of its supersets have the same support count~\cite{pasquier1999discovering}, where a set X is a superset of another set Y, if X contains all the elements of the set Y. For example, the itemsets \{A\}, \{C\}, \{D\} and \{C, D\} in Figure~\ref{fig:fpm} are not closed frequent patterns since their supersets \{A, B\}, \{B, C\}, \{B, D\} and \{B, C, D\} have the same support count. Therefore, \{B\}, \{A, B\}, \{A, D\}, \{B, C\}, \{B, D\} and \{B, C, D\} are closed frequent patterns since none of their supersets have the same support count. On the other hand, itemsets \{A, C\}, \{A, B, C\}, \{A, B, D\}, \{A, C, D\} and \{A, B, C, D\} have support counts of 1, 0, 1, 0 and 1, respectively, which is less than the minimum support count and thus are identified as non-frequent itemsets. Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we followed prior work~\cite{moradi2017quantifying, liu1999mining} and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences. For each pattern, the statistical significance of the difference between pattern occurrence in successful and unsuccessful communications was reported by computing \hl{the} p-value of Pearson's chi-square test.

@@

This study is part of a line of research to develop machine-learning models to annotate (code) and analyze patient-counselor communication patterns. We have previously reported on the development of probabilistic generative models~\cite{kotov2014towards, kotov2015interpretable} and application of novel features for maximum margin and deep learning classifiers~\cite{hasan2016study} with the goal of automated annotation of MI session transcripts. Experiments applying the annotation model to novel datasets are underway to assess the generalizability of the model to more diverse types of clinical encounters (e.g, email coaching to increase fruit and vegetable intake, HIV clinical care visits~\cite{carcone2018using}). We also developed and evaluated probabilistic and deep learning methods for the task of predicting \hl{the} success of motivational interviews at any point during the interview~\cite{hasan2018predicting}. In this paper, we present current work that builds on past work performed on automation of annotation, more specifically, the sequential analysis of patient-counselor communication data to identify the counselor communication strategies linked to the elicitation of change talk and sustain talk. Next steps include examining the performance of these models in diverse data sets representing different populations and behavioral problems. We are also experimenting with machine learning methods to segment (parse) the stream of communications in a healthy eating promotion intervention delivered by email into semantically coherent chunks (phrases, sentences or paragraphs), to which the annotation model can then assign a behavior code. Together, these models form the basis of a complete system to automatically code and analyze patient-counselor interactions. An automated system for behavioral coding and analysis could substantially accelerate the pace of research on the causal mechanisms of Motivational Interviewing and inform both the theory and clinical practice by providing clinicians with information about how to best tailor their communication strategies to different patient populations.
